D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-19 ootavacuous.tex
Credit where it is due for causality appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Convert CPPMEM litmus tests to herd7
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{listing}[tbp]
\begin{verbatim}
 1 int main() {
 2   atomic_int x = 2;
 3   volatile int r1;
 4   {{{ { x.store(42, memory_order_relaxed); // invented store
 5         x.store(3, memory_order_relaxed); }
 6   ||| r1 = x.load(memory_order_relaxed);
 7   }}};
 8   return 0;
 9 }
\end{verbatim}
\caption{CPPMEM Invented Store}
\label{lst:CPPMEM Invented Store}
\end{listing}

The reason that atomic stores cannot be invented is that doing so can
introduce new (and almost certainly undesirable) behaviors that are
forbidden by the abstract machine.
To see this, consider Listing~\ref{lst:CPPMEM Invented Store},
a CPPMEM\footnote{
	\url{http://svr-pes20-cppmem.cl.cam.ac.uk/cppmem/index.html}.}
litmus test that demonstrates such a behavior.
Without line~4, only the values~2 and~3 can be stored to \co{r1}.
With that line, the additional value~42 can also be stored to \co{r1}.
The implementation is therefore forbidden from inventing that store of 42
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
For example, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

@@

\begin{listing}[tbp]
\begin{verbatim}
 1 int main() {
 2   atomic_int x = 0;
 3   volatile int r1;
 4   {{{ { x.store(1, memory_order_relaxed); // duplicated store
 5         x.store(1, memory_order_relaxed); }
 6   ||| { r1 = x.load(memory_order_relaxed);
 7         x.store(2, memory_order_relaxed); }
 8   }}};
 9   return 0;
10 }
\end{verbatim}
\caption{CPPMEM Duplicated Store}
\label{lst:CPPMEM Duplicated Store}
\end{listing}

Duplicating atomic stores can also introduce new and undesirable
behaviors.
To see this, consider Listing~\ref{lst:CPPMEM Duplicated Store}, a CPPMEM
litmus test that demonstrates such a behavior.
Without line~4, if the final value of \co{r1} is 1, then the final value
of \co{x} must be 2.
With that line, the final value of \co{x} can be 1 even when the
final value of \co{r1} is 1.
The implementation is therefore forbidden from duplicating that store of 1
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
Again, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

@@

This Appendix is provided as a service to readers who want to know more
about the ongoing debates on causality.
References may be found here:

@@

% @@@ TODO:
% Convert remaining PPCMEM diagrams to herd7
% Emphasize that real-world implementations map abstract machine to matter
%	and energy in the real world.
% Look into why inkscape no longer likes generating PDFs.

>>>>

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/invented-store.litmus @@
\caption{Example Invented Store}
\label{lst:Example Invented Store}
\end{listing}

The reason that atomic stores cannot be invented is that doing so can
introduce new (and almost certainly undesirable) behaviors that are
forbidden by the abstract machine.
To see this, consider Listing~\ref{lst:Example Invented Store},
a \co{herd7} litmus test that demonstrates such a behavior.
Without line~8, only the values~0 and~3 can be loaded into \co{r1} on
line~13.
With line~8, the additional value~42 can also be loaded into \co{r1}.
The implementation is therefore forbidden from inventing that store of 42
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
For example, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

@@

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/duplicated-store.litmus @@
\caption{Example Duplicated Store}
\label{lst:Example Duplicated Store}
\end{listing}

Duplicating atomic stores can also introduce new and undesirable
behaviors.
To see this, consider Listing~\ref{lst:Example Duplicated Store}, a
litmus test that demonstrates such a behavior.
Without line~9, if the value loaded into \co{r1} is 1, then the final value
of \co{x} must be 2.
With line~9, the final value of \co{x} can be 1 even when the
value loaded into \co{r1} is 1.
The implementation is therefore forbidden from duplicating that store of 1
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
Again, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

@@

This Appendix is provided as a service to readers who want to know more
about the ongoing debates on causality, provided by the combined efforts
of Bard, Chat-GPT, and Bing Co-Pilot.
References may be found here:

@@

% @@@ TODO:
% Convert remaining PPCMEM diagrams to herd7
% Emphasize that real-world implementations map abstract machine to matter
%	and energy in the real world.
